---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 330,
  message = FALSE,
  warning = FALSE
)
```

```{r, eval = TRUE, include = FALSE}
require(emo)
```

# dPCRfit: Fit regression models to digital PCR data

<!-- badges: start -->
[![Lifecycle:
experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![MIT
license](https://img.shields.io/badge/License-MIT-blue.svg)](https://github.com/adrian-lison/dPCRfit/blob/main/LICENSE.md/)
[![DOI](https://zenodo.org/badge/947279049.svg)](https://zenodo.org/badge/latestdoi/947279049)
<!-- badges: end -->

## About
The `dPCRfit` R package allows fitting regression models to gene concentration
measurements obtained via digital PCR (dPCR). It uses a dPCR-specific likelihood
that accounts for non-detects and concentration-dependent measurement error (see
[here](https://doi.org/10.1101/2024.10.14.618307)). You can specify
regression models via a simple formula like in the base R `lm` function.

This package is useful for directly fitting regression models to dPCR
concentration measurements, for example when partition counts are not available.

Information about laboratory parameters (e.g. partition volume or sample
dilution) can be provided as priors. The package currently supports fixed-effect
linear models (identity or log link). Models are fitted using `stan` through the
`cmdstanr` interface, supporting exact estimates via MCMC sampling, or fast
approximation via pathfinder variational inference.

## Installing the package

Install the development version of `dPCRfit` from GitHub:
```{r, eval = FALSE}
remotes::install_github("adrian-lison/dPCRfit", dependencies = TRUE)
```

`r emo::ji("warning")` Note that `dPCRfit` also requires CmdStan to be installed on your system.
<details><summary>Click here for details on installing stan.</summary>

Simply run the `install_cmdstan()` function from `cmdstanr`. If you experience any problems installing CmdStan, see the [cmdstanr vignette](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) for help. 

```{r, eval = FALSE}
cmdstanr::check_cmdstan_toolchain()
cmdstanr::install_cmdstan(cores = 2) # use more cores to speed up
```

The stan model used by dPCRfit needs to be compiled for your device. This is only necessary once - after installing or updating the package - and can be done using the `dPCRfit_compile()` function.
```{r, eval = FALSE}
dPCRfit::dPCRfit_compile()
```

If the models are not successfully compiled, please ensure that `cmdstan` is properly set up and try updating it to a newer version using `cmdstanr::install_cmdstan()`. If the problem persists, please run `dPCRfit::dPCRfit_compile(verbose = TRUE)` and post the output in a new issue on GitHub, along with your `cmdstanr::cmdstan_version()`.
</details>

## Introduction
This short intro demonstrates how to fit a linear regression model to dPCR data using the `dPCRfit` package.
```{r}
library(dPCRfit) # load the dPCRfit package
library(ggplot2) # load ggplot2 for plotting
```

### Load example data
We load an example dataset with simulated dPCR measurements. The dataset
represents hypothetical measurements of a target gene concentration from samples
with different biomass.
```{r}
dPCR_data <- dPCR_linear_simulated
```

We can plot the simulated linear relationship between biomass and target gene
concentration using ggplot:
```{r data_plot}
ggplot(dPCR_data, aes(x=biomass, y=concentration)) +
  geom_point() +
  xlab("Biomass (mg)") + ylab("Target gene concentration [gc/mL]") +
  theme_bw()
```

## Fit a linear model
We will now use `dPCRfit` to fit a linear regression model to the simulated
concentration measurements. For this, we must first define a model of the
measurement noise. In our simulated data, we assumed a dPCR assay with the
following parameters:

- total number of partitions: 25'000
- concentration conversion factor : 1.73e-5 (gc/partition for 1 concentration unit in original sample)
- pre-PCR coefficient of variation: 10%

In our model, we will assume that we don't know the details of the dPCR assay. We therefore specify uncertain priors for the assay characteristics using the helper function `noise_dPCR()`:
```{r}
noise_model <- noise_dPCR(
          max_partitions_prior_lower = 10000, # partition number of chip between 10'000 and 30'000
          max_partitions_prior_upper = 30000,
          partition_loss_mean_prior_lower = 0.05, # expected share of invalid partitions between 5% and 30%
          partition_loss_mean_prior_upper = 0.3,
          partition_loss_variation_prior_lower = 0.1, # broad prior on variation of share of invalid partitions
          partition_loss_variation_prior_upper = 1,
          partition_loss_max = 0.5, # maximum share of invalid partitions is 50%
          volume_scaled_prior_mu = 1e-5, # broad prior for the concentration conversion factor
          volume_scaled_prior_sigma = 4e-5,
          prePCR_noise_type = "lognormal", # we assume that pre-PCR noise is log-normally distributed
        )
```
To summarize, we have made the following assumptions:

- we don't know what chip was used, but expect it to have between 10'000 and 30'000 partitions
- the average share of invalid partitions per dPCR run could be somewhere between 5% and 30%
- we don't know how much the share of invalid partitions varies between dPCR runs, so we use a broad prior for it
- we assume that PCR runs with more than 50% invalid partitions are discarded during quality control
- we think that the conversion factor is on the order of magnitude of 1e-5 but we use a broad prior for it

Using these priors specified in `noise_dPCR()`, the true values of these parameters will be estimated jointly from the data.

Once we have a suitable noise model, we can run a linear regression using `dPCRfit`:
```{r, cache = TRUE}
fitted_model <- dPCRfit(
        formula = concentration ~ biomass, # linear regression formula
        data = dPCR_data, # our data
        prior_intercept = c(0,1), # standard normal prior for the intercept
        prior_coefficients = c(0,1), # standard normal prior for the regression coefficients
        measurements = concentration_measurements(
          id_col = "sample_id", # name of the column with the sample id
          replicate_col = "replicate_id", # name of the column with the biological replicate id
          concentration_col = "concentration", # name of the column with the concentration measurements
          n_averaged_col = "n_technical_reps" # name of the column with the number of technical replicates per sample
        ),
        noise = noise_model, # the dPCR noise model we defined above
        nondetect = nondetect_dPCR() # also model non-detects from dPCR
      )
```

## Results
We can inspect our fitted model using the `summary()` function.
```{r}
summary(fitted_model)
```
Let's plot the fitted model against the data.
```{r model_fit_plot}
model_prediction <- predict(
  fitted_model,
  newdata = data.frame(biomass = seq(0,30)),
  interval = "confidence",
  keep_data = TRUE
  )

ggplot(dPCR_data, aes(x=biomass)) +
  geom_point(aes(y=concentration)) +
  geom_line(data = model_prediction, aes(x = biomass, y = mean), color = "blue") +
  geom_ribbon(data = model_prediction, aes(x = biomass, ymin = q5, ymax = q95), fill = "blue", alpha = 0.3) +
  xlab("Biomass (mg)") + ylab("Target gene concentration [gc/mL]") +
  theme_bw()
```

We can also print the fitted coefficients with 95% credible intervals. The dashed line shows the simulated ground truth.
```{r effects_plot}
ggplot(fitted_model$coef_summary) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(aes(y = variable, x = median)) +
  geom_errorbar(aes(y = variable, xmin = q5, xmax = q95), width = 0.2) +
  geom_vline(data = data.frame(variable = c("(Intercept)", "biomass"), true = c(0.2, 0.6)), aes(xintercept = true), linetype = "dotted", color = "darkred") +
  ylab("Coefficient") + xlab("Estimate") + facet_wrap(~variable, scales = "free", ncol = 1) +
  scale_x_continuous(expand = c(0.2, 0.2)) +
  theme_bw()
```
We see that the intercept was estimated with large uncertainty, due to the
increased noise at low concentrations. In contrast, the slope of the linear
relationship with biomass was estimated quite precisely.

## Citing the package

If you are using `dPCRfit` in a publication, please cite:

```
@article{lisonImprovingInferenceEnvironmental2025,
  title = {Improving inference in environmental surveillance by modeling the statistical features of digital {{PCR}}},
  author = {Lison, Adrian and Julian, Timothy and Stadler, Tanja},
  year = {2025},
  journal = {bioRxiv preprint},
  doi = {10.1101/2024.10.14.618307}
}
```
